%%
% 摘要信息
% OpsEvolver 档中前缀"c-"代表中文版字段, 前缀"e-"代表英文版字段
% 摘要内容应概括地反映出本论文的主要内容，主要说明本论文的研究目的、内容、方法、成果和结论。要突出本论文的创造性成果或新见解，不要与引言相 混淆。语言力求精练、准确，以 300—500 字为宜。
% 在摘要的下方另起一行，注明OpsEvolver 的关键词（3—5 个）。关键词是供检索用的主题词条，应采用能覆盖论文主要内容的通用技术词条(参照相应的技术术语 标准)。按词条的外延层次排列（外延大的排在前面）。摘要与关键词应在同一页。
% modifier: 黄俊杰(huangjj27, 349373001dc@gmail.com)
% update date: 2017-04-15
%%

\cabstract{
    随着云原生技术和微服务架构的普及，现代 IT 系统的复杂性与动态性显著增加，导致传统运维（Ops）面临巨大挑战。站点可靠性工程（SRE）通过将软件工程原则应用于运维实践，旨在提升系统可靠性，但其高度依赖人工分析和经验，难以适应快速变化的云环境。近年来，人工智能运维（AIOps）应运而生，试图利用机器学习和大数据技术自动化运维任务。然而，多数 AIOps 方案侧重于孤立的问题解决，难以实现端到端的自主决策和持续学习。

    为解决上述挑战，本文提出并实现了一套名为 OpsEvolver 的多智能体自主演化 SRE 框架，该框架融合了大语言模型（LLM）的推理能力与混沌工程实践，通过离线演化主动探索未知故障空间，并把诊断与反思过程中形成的经验持续整理为可复用于在线故障诊断的知识和优化故障诊断的上下文。其中多智能体协作体系由探索者、诊断者和反思者组成，探索者负责施加受控扰动以模拟真实故障，诊断者负责围绕告警组织根因分析，反思者则负责从诊断经验中提取结构化知识。

    OpsEvolver 在 Astronomy Shop、Hotel Reservation 和 Social Network 三个微服务系统上，分别基于 DeepSeek-V3.2、Llama 4 Maverick 17B Instruct 和 Qwen3-32B 累计执行了上千次故障注入，其中大部分故障触发了可观测告警并进入诊断流程，沉淀出丰富的知识条目和规则。实验结果表明，OpsEvolver 能够在真实微服务环境中组织完整的“演练-诊断-反思-沉淀”闭环，并显式学习可观测性边界：Astronomy Shop 更容易暴露时延传播型异常，Hotel Reservation 与 Social Network 则更突出日志主导和级联故障主导的排障模式。最后，本文从 AIOpsLab 问题集中选出了 Astronomy Shop 等三个微服务系统为背景的故障问题作为基准评测，成功验证了演化得到的知识能够改善故障根因分析（RCA）表现。尽管严格根因命中率仍有提升空间，OpsEvolver 已经证明，借助外部知识记忆、多模态告警建模和演练驱动演化，可以持续积累环境特定的运维知识，为构建具备自主演化能力的下一代智能运维系统提供可行路径。
}
% 中文关键词(每个关键词之间用“，”分开,最后一个关键词不打标点符号。)
\ckeywords{智能运维，大语言模型，智能体，站点可靠性工程，混沌工程，自适应}

\eabstract{
    With the proliferation of cloud-native technologies and microservice architectures, the complexity and dynamism of modern IT systems have increased significantly, posing serious challenges to traditional operations (Ops). Site Reliability Engineering (SRE) applies software engineering principles to operations in order to improve system reliability, but it still relies heavily on manual analysis and human experience, making it difficult to keep up with rapidly evolving cloud environments. In recent years, Artificial Intelligence for IT Operations (AIOps) has emerged as an attempt to automate operational tasks with machine learning and data-driven techniques. However, most existing AIOps approaches focus on isolated problems and therefore struggle to support end-to-end autonomous decision-making and continuous learning.

    To address these challenges, this thesis proposes and implements OpsEvolver, a multi-agent self-evolving SRE framework. By combining the reasoning ability of large language models (LLMs) with chaos engineering practice, OpsEvolver actively explores unknown fault spaces through offline evolution and continuously organizes the experience obtained from diagnosis and reflection into reusable knowledge for future fault diagnosis and optimize the diagnostic context. Its multi-agent collaboration mechanism consists of an explorer, a diagnoser, and an introspector. The explorer applies controlled perturbations to simulate realistic failures, the diagnoser performs root cause analysis around the observed alerts, and the introspector extracts structured knowledge from previous diagnostic experience.

    Across three representative microservice systems, namely Astronomy Shop, Hotel Reservation, and Social Network, OpsEvolver conducted more than a thousand fault injections based on DeepSeek-V3.2, Llama 4 Maverick 17B Instruct, and Qwen3-32B. Most of these injections triggered observable alerts and entered the diagnosis workflow, producing a substantial number of knowledge items and rules. The experimental results show that OpsEvolver can organize a complete loop of rehearsal, diagnosis, reflection, and knowledge accumulation in real microservice environments, while also learning observability boundaries explicitly. Astronomy Shop more readily exposes latency-propagation anomalies, whereas Hotel Reservation and Social Network reveal more log-dominant and cascading-failure-oriented troubleshooting patterns. Finally, this thesis constructs an independent benchmark from AIOpsLab problems based on the same three microservice systems, and the results further confirm that the knowledge accumulated through evolution can improve root cause analysis (RCA) performance. Although the strict root-cause hit rate still leaves room for improvement, OpsEvolver demonstrates that external knowledge memory, multi-modal alert modeling, and rehearsal-driven evolution can continuously accumulate environment-specific operational knowledge, providing a feasible path toward next-generation intelligent operations systems with self-evolving capability.
}
% 英文文关键词(每个关键词之间用,分开, 最后一个关键词不打标点符号。)
\ekeywords{AIOps, Large Language Model, Agent, Site Reliability Engineering, Chaos Engineering, Self-Adaptive}
